Ownership types in theory and practice (in Rust)

### 2023-01-17 paper/paper.tex
Improved figures

The core concept of ownership types is that objects can be \emph{owners} of other objects.
Given objects \emph{a} and \emph{b}, \emph{a} is \emph{inside} \emph{b} if \emph{a} is the same object as \emph{b} or \emph{a} is \emph{owned} by \emph{b}, transitively.
The \emph{outside} relation is the converse of the \emph{inside} relation.

In most ownership-type variants, the heap is divided into different regions called \empg{ownership contexts}.
Each object has one single ownership context, which is the set of all objects it owns.
Objects that are in the same ownership context are called \emph{siblings} or \emph{peers}.

@@

% \begin{figure}
%   \begin{tikzpicture}[object/.style={rectangle, draw, very thick, minimum size=20}]
%     \node[object] (A) {a};
%     \node[object] (C) [right=of A,yshift=1cm] {c};
%     \node[object] (B) [below=of A] {b};
%     \node[draw,thin,inner sep=2mm,fit=(A.west) (B) (B) (A.west)] {};

%     \draw[->,thick] (A.south) to node[left] {owns} (B.north);
%     \draw[->,thick,dashed] (C.south) to node[right] {no access} (B.north);
%   \end{tikzpicture}

%   \caption{Owners-as-modifiers \cite{ownership-types-survey}}
% \end{figure}

@@

    \draw[->,thick] (World.south) to node[left] {owns} (A.north);
    \draw[->,thick] (World.south) to node[right] {owns} (C.north);
    \draw[->,thick] (A.south) to node[left] {owns} (B.north);
    \draw[->,thick,dashed] (C.south) to node[right] {no access} (B.north);
  \end{tikzpicture}

  \caption{Owners-as-modifiers}
\end{figure}

An implication of this model is that any external reference to an object is only possible through its owner.
% TODO: what does this implicate further?

\paragraph{Advantages and disadvantages}

@@

@@

% TODO: is it obvious that this contains exact sentences from citation?
The paper that first introduced the idea of \emph{owners-as-ombudsmen} \cite{owners-as-ombudsmen} describes the concept as an extension of the owners-as-dominators model.
With owners-as-dominators, every path from the root to an object must pass through its owner.
Thus, encapsulated objects must have a single \emph{bridge object} that mediates all external interaction with its internal objects.
The owners-as-ombudsmen approach, on the other hand, relaxes the single bridge object constraint and allows several bridge objects to collectively define an aggregate with a shared representation.
These bridge objects are called \emph{ombudsmen} to emphasize their benevolent nature.

>>>>

The core concept of ownership types is that objects can be \emph{owners} of other objects.
Given the two objects \emph{a} and \emph{b}, \emph{a} is \emph{inside} \emph{b} if \emph{a} is the same object as \emph{b} or \emph{a} is \emph{owned} by \emph{b}, transitively.
Given the two objects \emph{a} and \emph{b}, \emph{a} is \emph{outside} \emph{b} if \emph{b} is the same object as \emph{a} or \emph{b} is \emph{owned} by \emph{a}, transitively.
The \emph{outside} relation is the converse of the \emph{inside} relation.

In most ownership-type variants, the heap is divided into different regions called \emph{ownership contexts}.
Each object has one single ownership context, which is the set of all objects it owns.
Objects that are in the same ownership context are called \emph{siblings} or \emph{peers}.

@@

@@

    \draw[->,thick] (World) to node[left] {owns} (A);
    \draw[->,thick] (World) to node[right] {owns} (C);
    \draw[->,thick] (A) to node[left] {owns} (B);
    \draw[->,thick,dashed] (C) to node[right] {invalid access} (B);
  \end{tikzpicture}

  \caption{Owners-as-modifiers}
  \label{fig:owners-as-modifiers}
\end{figure}

\begin{figure}
  \begin{tikzpicture}[object/.style={rectangle, draw, very thick, minimum size=20}]
    \node[object] (World) {world};
    \node[object] (Car) [below=of World] {car};
    \node[object] (Driver) [right=of Car,xshift=1.5cm] {driver};
    \node[object] (Engine) [below=of Car] {engine};

    \draw[->,thick] (World) to (Car);
    \draw[->,thick] (World) to (Driver);
    \draw[->,thick] (Car) to (Engine);
    \draw[->,thick,dashed] (Car) to node[above] {valid} (Driver);
    \draw[->,thick,dashed] (World) edge[bend right=45] node[left] {invalid} (Engine);
  \end{tikzpicture}

  \caption{Owners-as-modifiers \cite{flexible-alias-protection}, solid lines indicate "owns"}
  \label{fig:owners-as-modifiers2}
\end{figure}

% TODO: Add rust section on taking ownership of fields inside type
% Struct Car {
%   engine: Engine,
% }

% let car = Car::new(Engine::new());
% let driver = Driver::new();

% Abstract. Ownership types were devised nearly 15 years ago to provide
% a stronger notion of protection to object-oriented programming languages.
% Rather than simply protecting the fields of an object from external ac-
% cess, ownership types protect also the objects stored in the fields, thereby
% enabling an object to claim (exclusive) ownership of and access to other
% objects. Furthermore, this notion is statically enforced by now-standard
% type-checking techniques.
% Originating as the formalisation of the core of Flexible Alias Protec-

An implication of this model is that any external reference to an object is only possible through its owner.
Figure \ref{fig:owners-as-modifiers} provides an example of an invalid reference from object \emph{c} to \emph{b}, as there exists no direct link in the ownership hierarchy.
% TODO: what does this implicate further?

@@

% IMPORTANT TODO: question: can an object access owned children transitively?

% Advantages: useful for reasoning about properties of code.

@@

% TODO: is it obvious that this contains exact sentences from citation?
The paper that first introduced the idea of \emph{owners-as-ombudsmen} \cite{owners-as-ombudsmen} describes the concept as an extension of the owners-as-dominators model.
% TODO: this is the same as dominators
With owners-as-dominators, every path from the root to a given object must pass through that object's owner.
Thus, encapsulated objects must have a single \emph{bridge object} that mediates all external interaction with its internal objects.
The owners-as-ombudsmen approach, on the other hand, relaxes the single bridge object constraint and allows several bridge objects to collectively define an aggregate with a shared representation.
These bridge objects are called \emph{ombudsmen} to emphasize their benevolent nature.
